---
title: "Normali komanda - humanMethylation"
output: html_document
---

# Setup

```{r setup, include=FALSE, out.width=792}

knitr::opts_chunk$set(echo = TRUE)

library(FlowSorted.Blood.450k)
library(minfi)
library(IlluminaHumanMethylation450kmanifest)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(cluster)
library(knitr)
library(kableExtra)

```

# 1. Clean up the annotation
# 2. Prepare and save the RGChannelSet

Due to the low quality of previous data, we will be using already prepared raw data objects from Bioconductor experiment package FlowSorted.Blood.450k.

A few samples from the dataset:

```{r prepareData}

# Load data
data <- FlowSorted.Blood.450k
saveRDS(data, file="../../output/step_1/RGSet.RDS")

# Show a few samples information
samples_list <- head(pData(data), 10)
kable(samples_list, "html") %>%
      kable_styling(font_size=10) %>% 
      kable_styling("striped") %>%
      scroll_box(width="100%")

# Remove redundant variables
rm(samples_list)

```

# 3. Obtain detection p-values

All values with detection p-value > 0.01 are considered as "bad". As we can see from the following code, there are no samples that have more than 1% of "bad" values. Therefore, no samples have been removed in this step.

The graph below shows the first 10 samples p-value percentage. Small values indicate that the dataset is of a high quality.

```{r pValues}

# Obtain detection p-values
detP <- detectionP(data)

# All values with detection p-value > 0.01 are "bad"
failed <- detP > 0.01

# Good sample = fraction of failed positions < 0.01
goodSamples <- colMeans(failed) < 0.01
badSamples  <- colMeans(failed) >= 0.01

# Count good and bad samples
count(goodSamples)
count(badSamples)

# Barplot showing first 10 samples p-value percentage
randomSamples <- head(colMeans(failed), 10)

b <- barplot(
  horiz=TRUE,
  randomSamples,
  las=2,
  col=c("darkviolet", "darkturquoise"),
  main="Percentage of failed positions"
)

text(
  y=b, 
  x=randomSamples-0.00005, 
  labels=as.character(format(round(randomSamples, 7), nsmall = 2))
)

# Remove redundant variables
rm(b)
rm(goodSamples)
rm(badSamples)
rm(randomSamples)

```

# 4. Normalize the dataset

```{r normalize}

data <- preprocessSWAN(data)

```

# 5. Remove all probes that have more than 1% of “bad” detection p-values

```{r filterProbes}

# Remove all probes that are "bad"
goodProbes <- rowMeans(failed) < 0.01
data <- data[goodProbes, ]

# Remove redundant variables
rm(failed)
rm(goodProbes)

```

# 6. Remove “CH” probes and “SNP related” probes

```{r removeCH}

data <- dropMethylationLoci(data)

```

# 7. Extract the 3 tables from the prepared object file

```{r extractTables}

# The main matrix of values
dataMatrix <- getBeta(data)

# The sample key information
keyInfo <- pData(data)

# The meta information about rows
metaInfo <- getAnnotation(data)

```

# 8. Implement the IAC outlier removal method

The first three graphs represent the original dataset without outlier removal.
We've chosen sdout = -2 and performed two iterations to elimiate outliers.
The following graphs represent the quality increase with each iteration.

```{r outlierRemoval}

# Calculating IACs for all pairs of samples and
# examining the distribution of IACs in the dataset
dim(dataMatrix)
IAC <- cor(dataMatrix, use="p")
hist(
  IAC,
  sub=paste("Mean=", format(mean(IAC[upper.tri(IAC)]), digits=3)),
  col=c("darkviolet", "darkturquoise")
)

# Performing hierachical clustering using 1-IAC as a distance metric
cluster <- hclust(as.dist(1-IAC), method="average")
plot(
  cluster,
  cex=0.7,
  labels=dimnames(dataMatrix)[[2]]
)

# Another way to visualize outliers 
# Calculating the mean IAC for each array
meanIAC <- apply(IAC, 2, mean)
sdCorr <- sd(meanIAC)
numbersd <- (meanIAC-mean(meanIAC))/sdCorr
plot(numbersd)
abline(h=-2)

# Outliers depicted above are the 
# same outliers identified in the initial dendrogram
sdout <- -2
outliers <- dimnames(dataMatrix)[[2]][numbersd<sdout]
outliers

# Removing outliers and calculating new IAC matrix
dataMatrix2 <- dataMatrix[, numbersd>sdout]
dim(dataMatrix2)

###
IAC <- cor(dataMatrix2, use="p")
hist(
  IAC,
  sub=paste("Mean=", format(mean(IAC[upper.tri(IAC)]), digits=3)),
  col=c("darkviolet", "darkturquoise")
)

###
cluster <- hclust(as.dist(1-IAC), method="average")
plot(
  cluster,
  cex=0.7,
  labels=dimnames(dataMatrix2)[[2]]
)

###
meanIAC <- apply(IAC, 2, mean)
sdCorr <- sd(meanIAC)
numbersd <- (meanIAC-mean(meanIAC))/sdCorr
plot(numbersd)
abline(h=-2)

# Repeating the same process again
sdout <- -2
outliers <- dimnames(dataMatrix2)[[2]][numbersd<sdout]
outliers

###
dataMatrix3 <- dataMatrix2[, numbersd>sdout]
dim(dataMatrix3)

###
IAC <- cor(dataMatrix3, use="p")
hist(
  IAC,
  sub=paste("Mean=", format(mean(IAC[upper.tri(IAC)]), digits=3)),
  col=c("darkviolet", "darkturquoise")
)

###
cluster <- hclust(as.dist(1-IAC), method="average")
plot(
  cluster,
  cex=0.7,
  labels=dimnames(dataMatrix3)[[2]]
)

###
meanIAC <- apply(IAC, 2, mean)
sdCorr <- sd(meanIAC)
numbersd <- (meanIAC-mean(meanIAC))/sdCorr
plot(numbersd)
abline(h=-2)

# Removing redundant variables
rm(IAC)
rm(cluster)
rm(dataMatrix2)
rm(dataMatrix3)
rm(outliers)
rm(meanIAC)
rm(sdCorr)
rm(numbersd)
rm(sdout)

```

# 9. Add a short quality control step

For quality control step, methylation levels at different CpG Island positions were compared.
As we can see from the graph below, seas and shelves have high methylation rates, whereas islands and shores have low methylation.

``` {r qualityControl}

# Comparing methylation levels at different CpG Island positions
cpgMeans <- rowMeans(dataMatrix)

names(cpgMeans) <- metaInfo$Relation_to_Island

shore   <- cpgMeans[names(cpgMeans) %in% c("N_Shore", "S_Shore")]
shelf   <- cpgMeans[names(cpgMeans) %in% c("N_Shelf", "S_Shelf")]
island  <- cpgMeans[names(cpgMeans) == "Island"]
sea     <- cpgMeans[names(cpgMeans) == "OpenSea"]

plot(
  density(shore),
  ylim=c(0,8),
  col="deeppink",
  main="Methylation levels at different CpG Island positions",
  cex=2
)

lines(density(shelf), col="darkviolet", cex=2)
lines(density(island), col="darkturquoise", cex=2)
lines(density(sea), col="green", cex=2)

legend(
  x=0.4,
  y=8,
  legend=c("Shore", "Shelf", "Island", "Sea"),
  col=c("deeppink", "darkviolet", "darkturquoise", "green"),
  lty=1
)

# Remove redundant variables
rm(cpgMeans)
rm(island)
rm(sea)
rm(shelf)
rm(shore)

```

# 10. Save the prepared data objects to the output directory

``` {r saveData}

saveRDS(data, "../../output/step_1/finalData.RDS")
saveRDS(dataMatrix, "../../output/step_1/dataMatrix.RDS")
saveRDS(keyInfo, "../../output/step_1/keyInfo.RDS")
saveRDS(metaInfo, "../../output/step_1/metaInfo.RDS")

```




